Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-20 manuscript/manuscript.tex
streamlined data section

\section{Introduction} 
\label{sec:intro}
In the early 21st century,
the open data movement began to take shape, driven by the
fundamental belief that freely accessible data can transform 
both societies and governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
in breaking barriers that once kept valuable information 
on government performance in silos. Their collective impact 
is profound, extending beyond mere data sharing to 
fostering a culture of openness that benefits individuals, 
communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

@@

The paper is organized as follows:
Section~\ref{sec:history} offers a brief review of the history of the 311 system. In 
Section~\ref{sec:trends} we take a look at long-term trends presented 
via a 10-year analysis. Section~\ref{sec:issues} offers a 
general discussion of data cleansing issues impacting data 
quality and curation efficiency. Section~\ref{sec:structural} examines
the technical dimensions of the dataset, the structural issues. 
In Section~\ref{sec:datatypes} we examine the data fields for compliance 
with the Data Dictionary. Section~\ref{sec:blanks} looks at the 
data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} 
explores how data fields comply with a domain of legal 
or acceptable values. Section~\ref{sec:inconsistencies} deals 
with the important issues surrounding logical inconsistencies 
and concerning patterns in the data. Section~\ref{sec:precision} 
explores the issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The 
paper concludes with a discussion and recommendations in 
Section~\ref{sec:discussion}, encapsulating the key insights and 
implications of our research.

@@

\section{NYC 311 System History} 
\label{sec:history}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. The evolution of this system can be traced from
its initial implementation as a simple inquiry channel via telephone to a
comprehensive data management system that handles millions of requests
annually. Each inquiry is logged as a complaint, a Service Request (SR), 
in the 311 system. Key milestones in the 311 system include:

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only. 
 	The ``311'' phone number replaces a myriad of numbers to 
 	call for each individual City Agency.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all 
   	Service Requests come from mobile \& online channels.
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 
      SRs hit an all-time high of 348,463. Seven of the Top 10 all-time busiest 
      days occur in 2020.
      \item 2021 - 311 System is expanded to include the MTA's city subway system, 
      the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

Today, the NYC 311 data system is a robust platform that manages a
vast array of urban living-related inquiries. The system handles over 3 million service 
requests (SRs) per year, encompassing such complaints as street noise, 
illegal parking, heat/hot water, abandoned vehicles,
and unsanitary conditions. The data is a valuable resource for City
administration and policy-making. The 311 data is publicly accessible through the 
NYC Open Data Portal which provides access to all the data 
sets (approx. 2700) and provides a convenient method for querying, 
grouping, aggregating, geo-mapping, visualization, and exporting 
results. This open data  enables not only governmental transparency 
but empowers researchers, civic developers, and the general public.

Despite its success, the 311 system faces challenges such as:

\begin{itemize}
	\item Data timeliness, accuracy, and consistency
	\item Difficulties correlating data over long time periods, e.g. over a 10-year 
	period, e.g. Agency name changes, new complaint types.
	\item Data anonymization and handling of Personally identifiable information (PII)
	\item Integration with stand-alone systems at selected NYC Agencies
	\item Managing API usage, authentication, and load for 3\textsuperscript{rd} 
	party users, of which there are many
\end{itemize}

These and other challenges are addressed by the various 
Agency open data managers as well as the 
\href{https://www.nyc.gov/content/oti/pages/}
{NYC Office of Technology and Innovation (OTI)} which provides the 
technology support for the open data system. Typically, the data 
will be exported and processed in an external program. The authors found 
this to be the case with the large 311 dataset examined.

The impact of the NYC 311 data extends beyond operational efficiency;
it has become instrumental in shaping City governance and community
engagement. The data has been pivotal in such areas as:

\begin{itemize}
	\item Providing advice on shelters during hurricanes, heat emergencies, and winter blizzards
	\item Handling countless inquires during the COVID pandemic 
	(testing centers, vaccine sites, mobile vaccine clinics, etc.)
	\item Enforcement of standards between landlords and tenants
	\item Re-allocation of NYC's yellow taxi routes based upon an insightful analysis 
	conducted by the Taxi \& Limousine Commission (TLC)
	\item Improved responsiveness for City Agencies with regards to direct 
	client-facing responsibilities such as Parks \& Rec, Dept of Transportation, DSNY, etc.
	\item Increased 311 SRs related to Homelessness resulted in the launch 
	of the City's Homeless Outreach \& Mobile Engagement Street Action Teams (HOME-STAT)
	which includes street canvassers and increases in homeless housing facilities.
\end{itemize}

We extracted 311 SR data for the period of CY 2022-2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.

\section{10-year Analysis \& Trends} 
\label{sec:trends}
As part of this study, we also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame . During that time-frame, 
the number of 311 SRs grew by 50\%. The summer and early fall months 
make up all of the Top 5 SR counts by day. The all-time daily 
high for SRs, 24,415 SRs (a full 9.8 \textsigma's from the 10-year mean), 
occurred on Tuesday, 2020-08-04. The 10-year low count occurs 
on Christmas Day, 2016, a Sunday.

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=\textwidth]{10-year-trend-monthly.pdf}
  	\caption{10-year (2014-2023) Monthly SR Counts}
  	\label{fig:10-yr-monthly}
\end{figure}

\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
	    \begin{tabular}{@{}lll lll@{}}
	        \toprule
	        \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
	        \cmidrule(r){1-2} \cmidrule(l){3-4}
	        Date & Count & Date & Count\\
	        \midrule
	        2020-08-04 & 24,415 & 2014-12-25 & 2965\\
	        2020-08-05 & 19,560 & 2015-12-25 & 3247\\
	        2023-09-29 & 17,962 & 2014-04-20 & 3287\\
	        2020-07-05 & 16,916 & 2015-12-26 & 3405\\
	        2020-06-21 & 15,883 & 2016-11-06 & 3419\\
	        \bottomrule
	    \end{tabular}
    \label{tab:top-10-counts}
\end{table}

Two important fields in this analysis effort are the responsible NYC Agency, 
and the type of complaint. Given the size of the City of New 
York government, it is necessary to identify the responsible 
Agency as well as the type of complaint in order to identify 
the responsible party and assist in troubleshooting discrepancies.
Thus we can often identify if it is an Agency specific problem, 
or a systemic issue. Although 311 SRs involve 16 different agencies, 
there are six Agencies that comprise 90\% of the complaints.

\begin{itemize}
	\item New York Police Department (NYPD)
	\item Housing Preservation \& Development (HPD
	\item New York City Department of Sanitation (DSNY)
	\item Department of Transportation (DOT)
	\item Department of Environmental Protection (DEP)
	\item Department of Parks \& Recreation (DPR)
\end{itemize}

\begin{figure}[tbp]
	\centering
	\includegraphics[width=\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

The distribution of complaint\_type skews to a few select complaints 
accompanied by a long tail. This graph shows the top 20 complaint\_type(s) 
comprising 70\% out of a total of 220 different types.

\begin{figure}[tbp]
	\centering
 	\includegraphics[width=\textwidth]{SR_by_Complaint_Type.pdf} 
	\caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
	\label{fig:SR_complaints}
\end{figure}

The top 20 complaints contain several of the eight different types of ``Noise'' 
complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total a full 22\% of all complaint types, which, when consolidated, is the most frequently occurring complaint\_type.

\begin{table}[tbp]
    \centering
    \caption{Noise-related complaints\_type(s) by count with Agency}
	    \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
	    round-precision=0]S[table-format=2.2,round-mode=places,
	    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
	        \toprule
	        \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
	        \midrule
	        Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
	        Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
	        Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
	        Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
	        Noise                      & 100413 &  1.57 & DEP   \\ 
	        Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
	        Noise - Park               &  16620 &  0.26 & NYPD  \\ 
	        Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
	        \bottomrule
	    \end{tabular}
    \label{tab:noisecomplaints}
\end{table}

>>>>

\section{Introduction} 
\label{sec:intro}

In the early 21st century,
the open data movement began to take shape, driven by the
fundamental belief that freely accessible data can transform 
both societies and governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
in breaking barriers that once kept valuable information 
on government performance in silos. Their collective impact 
is profound, extending beyond mere data sharing to 
fostering a culture of openness that benefits individuals, 
communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

@@

The paper is organized as follows:
Section~\ref{sec:data} offers a brief review of the history of the 311
system and the long-term trends presented via a 10-year analysis.
Section~\ref{sec:issues} offers a 
general discussion of data cleansing issues impacting data 
quality and curation efficiency. Section~\ref{sec:structural} examines
the technical dimensions of the dataset, the structural issues. 
In Section~\ref{sec:datatypes} we examine the data fields for compliance 
with the Data Dictionary. Section~\ref{sec:blanks} looks at the 
data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} 
explores how data fields comply with a domain of legal 
or acceptable values. Section~\ref{sec:inconsistencies} deals 
with the important issues surrounding logical inconsistencies 
and concerning patterns in the data. Section~\ref{sec:precision} 
explores the issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The 
paper concludes with a discussion and recommendations in 
Section~\ref{sec:discussion}, encapsulating the key insights and 
implications of our research.

@@

\section{NYC 311 Service Request Data} 
\label{sec:data}

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year, covering issues such as street noise, illegal
parking, heat and hot water, abandoned vehicles, and unsanitary
conditions. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

Despite its success, the 311 system faces several challenges, such as
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, handling of personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage and load for numerous
third-party users. These challenges are addressed by various agency
open data managers and the
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system. Data often needs to be exported and processed in
external programs to address these issues, as observed in the
examination of the large 311 dataset. Furthermore, the system must
adapt to changes such as agency name changes and the introduction of
new complaint types, which can complicate long-term data analysis.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
improving responsiveness for city agencies like Parks and Recreation,
the Department of Transportation, and the Department of Sanitation,
and addressing homelessness through initiatives like the Homeless
Outreach and Mobile Engagement Street Action Teams (HOME-STAT). These
applications demonstrate the transformative potential of open data in
enhancing urban living and fostering a more responsive and transparent
city government.

\jy{FixME: this paragraph's location will be revisited later.}
We extracted 311 SR data for the period of CY 2022--2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.

\jy{Every float (figure/table) needs to be referred to in the text.}

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend-monthly.pdf}
  	\caption{10-year (2014-2023) Monthly SR Counts}
  	\label{fig:10-yr-monthly}
\end{figure}

We also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by 50\%. The summer and early fall months 
make up all of the Top 5 SR counts by day.

\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
    \label{tab:top-10-counts}
    \begin{tabular}{@{}lll lll@{}}
      \toprule
      \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
      \cmidrule(r){1-2} \cmidrule(l){3-4}
      Date & Count & Date & Count\\
      \midrule
      2020-08-04 & 24,415 & 2014-12-25 & 2965\\
      2020-08-05 & 19,560 & 2015-12-25 & 3247\\
      2023-09-29 & 17,962 & 2014-04-20 & 3287\\
      2020-07-05 & 16,916 & 2015-12-26 & 3405\\
      2020-06-21 & 15,883 & 2016-11-06 & 3419\\
      \bottomrule
    \end{tabular}
\end{table}

Table~\ref{tab:top-10-counts} summarizes the top 5 and bottom 5 days
in counts of SRs. The highest daily SR count, 24,415, occurred on
August 4, 2020, coinciding with the aftermath of Tropical Storm
Isaias, highlighting the system's critical role during crises. The
second-highest count, 19,560, was the following day, reflecting the
storm's continued impact. 
Conversely, the lowest daily SR count, 2,965, was on Christmas Day,
2014, with similar lows on other Christmas Days, indicating reduced
activity during major holidays. The data also shows fewer requests on
weekends, such as April 20, 2014 (Easter Sunday), and November 6, 2016
(a non-holiday Sunday).

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

Figure~\ref{fig:SRcountbyAgency} display the SR counts by agency with
cumulative percentage. Two important fields in this analysis are the
responsible NYC Agency and the type of complaint. Given the size of
the City of New York government, identifying the responsible agency
and the type of complaint is essential for pinpointing the accountable
party and troubleshooting discrepancies. This identification helps
determine whether an issue is specific to an agency or indicative of a
systemic problem. Although 311 SRs involve 16 different agencies, six
agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}

The distribution of complaint types skews toward a few select
categories, accompanied by a long tail. Figure~\ref{fig:SR_complaints}
shows that the top 20 complaint types account for 70\% of the total
220 different types. Notably, these top complaints include several
variations of noise-related issues, which collectively represent 22\%
of all complaints. When consolidated, noise complaints emerge as the
most frequently occurring type. Table~\ref{tab:noisecomplaints}
provides a detailed breakdown of noise-related complaint types by
count and responsible agency. The most common noise complaint is
``Noise - Residential,'' with 675,502 complaints, accounting for 10.56\%
of all service requests, and primarily handled by the NYPD. Other
significant noise complaints include ``Noise - Street/Sidewalk"
(4.70\%), ``Noise - Commercial'' (2.02\%), and ``Noise - Vehicle"
(1.81\%), all managed by the NYPD. Additional categories such as
"Noise - Helicopter'' (1.33\%) fall under the jurisdiction of the EDC,
and ``Noise'' (1.57\%) is managed by the DEP. The prevalence of these
complaints underscores the critical role of the NYPD and other
agencies in addressing noise-related issues, reflecting the
significant impact of urban noise on residents' quality of life. This
data highlights the importance of efficient response and management
strategies to mitigate noise disturbances in New York City.
